---
title: "Admixture plots"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Admixture plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggPopi)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggPopi", mustWork = TRUE)
}
```

`ggpop` keeps the user-facing admixture API narrow:

- `import_admix()` creates typed `ggpop_admix` objects;
- `plot_admix()` returns a `ggplot` object;
- `ggpop() + geom_admix()` gives the extension-style layered workflow.

## API summary

| Task | API | Notes |
|---|---|---|
| Import ADMIXTURE directory or `.Q` files | `import_admix(file, type = "admixture", pop_group = NULL)` | Reads full K result sets |
| Import STRUCTURE-style numeric Q matrix | `import_admix(file, type = "structure")` | Limited numeric Q support |
| Direct plot | `plot_admix(data, k = ...)` | `k = "all"`, one K, or a vector |
| Layered plot | `ggpop(data) + geom_admix(k = ...)` | Main ggplot extension path |
| Original pophelper behavior | `plot_pophelper_q()` | Advanced compatibility layer |

## Full K results

```{r}
admix <- import_admix(
  ggpop_extdata("admixture"),
  type = "admixture",
  ind = ggpop_extdata("snp", "finalsnp_ld.fam"),
  pop_group = ggpop_extdata("pop_group.txt")
)
```

```{r, fig.alt = "Faceted stacked bar chart. Each panel represents one K value from the imported ADMIXTURE directory, individuals are arranged along the x-axis, and stacked bar segments show ancestry proportions that sum to one for each individual."}
plot_admix(admix, k = "all")
```

```{r, fig.alt = "Faceted stacked bar chart with two selected K panels. The K equals 2 and K equals 4 panels show how ancestry proportions split into more clusters when a higher K value is selected."}
plot_admix(admix, k = c(2, 4))
```

```{r, fig.alt = "Stacked bar chart. Individuals are on the x-axis and ancestry proportions are stacked within each bar for K equals 3, demonstrating that the ggplot layer path produces the same kind of admixture display as the direct plot."}
ggpop(admix) + geom_admix(k = 3)
```

## Population group labels and sorting

`pop_group.txt` uses two columns, `sample` and `pop`. It is joined during import:

```{r}
head(import_pop_group(ggpop_extdata("pop_group.txt")))
```

The native ggplot implementation mirrors the supported pophelper `plotQ()` behavior for:

- individual labels with `show_sample_labels = TRUE`;
- sorting individuals with `sort = "all"`, `sort = "label"`, or a cluster name such as `K1`;
- group labels with `show_group_labels = TRUE`;
- sorting with group labels using `order_group = TRUE`.

```{r, fig.alt = "Grouped stacked bar chart. Individuals are sorted within population groups for K equals 3, ancestry cluster proportions fill each bar, and population group strip labels separate the sample groups."}
plot_admix(
  admix,
  k = 3,
  sort = "all",
  order_group = TRUE,
  show_group_labels = TRUE,
  show_sample_labels = TRUE
)
```

The layered ggplot path returns the same kind of ggplot object:

```{r, fig.alt = "Layered grouped stacked bar chart. Individuals are ordered by ancestry cluster and population group, showing that the geom_admix layer preserves the same sorting and grouping controls as plot_admix."}
ggpop(admix) +
  geom_admix(k = 3, sort = "all", order_group = TRUE)
```

## Direct vs layered use

`plot_admix()` is the direct plot wrapper.
`geom_admix()` is the layered interface when you want to build a larger `ggplot` object yourself.
Both routes use the same visual defaults; `plot_admix()` is the reference style,
and `ggpop(admix) + geom_admix()` reproduces that publication-level look inside a ggplot composition.

```{r, fig.alt = c("Stacked bar chart from plot_admix. Individuals are arranged along the x-axis and ancestry proportions are stacked within each bar for K equals 3.", "Stacked bar chart from ggpop plus geom_admix. The same K equals 3 ancestry proportions are displayed through the layered ggplot extension path.")}
plot_admix(admix, k = 3)
ggpop(admix) + geom_admix(k = 3)
```

## STRUCTURE-style input

`import_admix()` also accepts limited STRUCTURE-style inputs.
The importer always returns the same typed long-format object.

```{r eval=FALSE}
structure <- import_admix("structure_results.out", type = "structure")
plot_admix(structure, k = "all")
```
