---
title: "Compatibility and advanced helpers"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Compatibility and advanced helpers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggPopi)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggPopi", mustWork = TRUE)
}
```

The recommended user-facing interfaces are:

- `import_gwas()` |> `plot_manha()` or `ggpop()` + `geom_manha()`
- `import_gwas()` |> `plot_qq()` or `ggpop()` + `ggpop::geom_qq()`
- `plot_admix()`
- `ggpop() + geom_admix()`

The package also exports advanced helpers for users who need original-package
compatibility or integration with older code.

## pophelper compatibility

```{r eval=FALSE}
plot_pophelper_q(import_admix("runs/", type = "admixture"), returnplot = TRUE)
plot_admixture_pophelper(import_admix("runs/", type = "admixture"))
```

These helpers keep the original `pophelper` return structure.

## Importing pophelper qlists

```{r eval=FALSE}
read_pophelper_q(files = "run1.Q", as_ggpop = TRUE)
```

When `as_ggpop = TRUE`, compatibility input is converted to a typed `ggpop_admix`
object so it can enter the same plotting pipeline as native imports.
