---
title: "GWAS plots"
output:
  rmarkdown::html_vignette:
    fig_width: 6.5
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{GWAS plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggpop)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggpop", mustWork = TRUE)
}
```

`ggpop` supports Manhattan and Q-Q plots from common GWAS outputs.
The intended flow is:

1. import a typed object;
2. plot with `plot_manha()` / `plot_qq()` or `ggpop() + geom_*()`;
3. keep [`fastman`](https://github.com/adhikari-statgen-lab/fastman) as the optional fast path when installed.

## API summary

| Task | Direct API | Layer API | Notes |
|---|---|---|---|
| Import GCTA/GEMMA/EMMAX | `import_gwas(file, type = ...)` | - | Returns `ggpop_gwas` |
| Manhattan | `plot_manha()` | `ggpop(data) + geom_manha()` | Defaults are aligned |
| Q-Q | `plot_qq()` | `ggpop(data) + ggpop::geom_qq()` | Use explicit namespace to avoid `ggplot2::geom_qq()` |
| Backend | `use_fastman = TRUE` | native ggpop stat | Direct wrappers use fastman when available |

## GCTA example

```{r}
gwas <- import_gwas(ggpop_extdata("gwas", "gcta.mlma"), type = "gcta")
```

```{r, fig.alt = "Manhattan plot. Chromosomes are arranged along the x-axis and minus log10 p-values are on the y-axis, with threshold lines marking suggestive and genome-wide significance."}
plot_manha(gwas, title = "GCTA Manhattan", use_fastman = TRUE)
```

```{r, fig.alt = "Q-Q scatter plot. Expected minus log10 p-values are on the x-axis and observed minus log10 p-values are on the y-axis, with a diagonal reference line and genomic inflation label."}
plot_qq(gwas, title = "GCTA Q-Q", use_fastman = TRUE)
```

## Layered workflow

```{r, fig.alt = "Layered GWAS examples showing a Manhattan plot followed by a Q-Q scatter plot using the same imported GWAS object."}
gwas |>
  ggpop() +
  geom_manha()

gwas |>
  ggpop() +
  ggpop::geom_qq()
```

`plot_manha(gwas, use_fastman = FALSE)` and `ggpop(gwas) + geom_manha()`
share the same default threshold and suggestive reference lines.

## GWAS colour palettes

The GWAS module uses the same unified discrete palette entry as the rest of
`ggpop`. Manhattan plots default to one discrete colour per chromosome:

```{r, fig.alt = "Manhattan plot using the default ggpop manhattan palette with one discrete colour assigned per chromosome."}
plot_manha(gwas, use_fastman = FALSE, palette = "manhattan")
```

For a two-colour alternating Manhattan plot, set `binary = TRUE` and pass either
a ggpop palette name or two explicit colours:

```{r, fig.alt = "Manhattan plot using a two-colour alternating palette repeated across chromosomes."}
plot_manha(
  gwas,
  use_fastman = FALSE,
  palette = c("#123456", "#654321"),
  binary = TRUE
)
```

## Alternative GWAS formats

`import_gwas()` accepts `type = "gcta"`, `type = "gemma"`, and `type = "emmax"`.
The package also keeps typo-compatible aliases for old prompts and notebooks.
They are compatibility helpers, not recommended in new code:

```{r eval=FALSE}
improt_gwas("assoc.mlma", type = "gcta")
prot_gwas("assoc.mlma", type = "gcta")
```

If `fastman` is installed, the direct GWAS wrappers use it by default.
If it is not installed, the ggplot-native path still works.
