---
title: "Introgression"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.8
vignette: >
  %\VignetteIndexEntry{Introgression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggPopi)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggPopi", mustWork = TRUE)
}
```

`ggpop` imports common introgression summaries into a typed
`ggpop_introgression` object. The module covers window statistics from Dsuite
and genomics_general, trio-level Dsuite D-statistics, and edge tables from
TreeMix-style migration summaries or ADMIXTOOLS2 qpGraph outputs.

## API summary

| Task | API | Notes |
|---|---|---|
| Import Dsuite Dtrios | `import_introgression(file, type = "dsuite_dtrios")` | Trio-level D statistic summaries |
| Import Dsuite local windows | `import_introgression(file, type = "dsuite_dinvestigate")` | D, fd, fdM, and df window statistics |
| Import genomics_general windows | `import_introgression(file, type = "genomics_general")` | ABBA-BABA window outputs |
| Import graph edges | `import_introgression(file, type = "treemix")` / `"qpgraph"` | Reads user-facing `from` / `to` edge tables |
| Direct plot | `plot_introgression(data, stat = ...)` | Returns a `ggplot` object |
| Layered plot | `ggpop(data) + geom_introgression(...)` | Tidy ggplot extension path |

## Window introgression statistics

genomics_general ABBA-BABA windows are imported in long form, with one row per
statistic per window. Genome-wide calls default to a chromosome-wise window
point plot on a Manhattan-like genome axis, matching common ABBA-BABA/fdM
window summaries where each point is one genomic window.

```{r}
intro <- import_introgression(
  ggpop_extdata("introgression", "genomics_general"),
  type = "genomics_general"
)
class(intro)
unique(intro$stat)
```

```{r, fig.alt = "Window introgression plot. D and fdM statistics are shown as chromosome-wise Manhattan-like points in stacked panels over chromosomes 1 and 2."}
plot_introgression(intro, stat = c("D", "fdM"))
```

When a chromosome or region is supplied, the direct plot uses a local position
axis in megabases.

```{r, fig.alt = "Regional introgression plot. D statistic windows on chromosome 1 are shown as a point-and-line trace over genomic position in megabases."}
plot_introgression(intro, stat = "D", chr = "1")
```

Dsuite `Dinvestigate` / localFstats files use the same plotting path.

```{r}
dsuite_windows <- import_introgression(
  ggpop_extdata("introgression", "Dsuite", "Dinvestigate.tsv"),
  type = "dsuite_dinvestigate"
)
unique(dsuite_windows$stat)
```

## Trio D-statistics

Dsuite `Dtrios` summaries are imported as trio-level statistics. The default
plot orders trios by D statistic value and keeps the direction visible.

```{r, fig.alt = "Trio-level D statistic dot plot. Population trios are listed on the y-axis and D statistic values are shown on the x-axis relative to zero."}
dtrios <- import_introgression(
  ggpop_extdata("introgression", "Dsuite", "Dtrios.tsv"),
  type = "dsuite_dtrios"
)
plot_introgression(dtrios)
```

## Graph edge tables

TreeMix and qpGraph workflows often need graph visualization after model
fitting. `ggpop` reads a compact edge table with `from` and `to` columns, plus
optional weights or bounds. TreeMix internal `*.edges.gz` files are not treated
as a stable public schema; convert migration edges from the user-facing
`treeout.gz` summary or provide an explicit edge table. Full TreeMix-style
`plot_tree()` plus residual heatmap support belongs to a dedicated TreeMix view.

```{r, fig.alt = "Introgression graph edge diagram. Root, source, and target population nodes are connected by directed migration or graph edges."}
treemix_edges <- import_introgression(
  ggpop_extdata("introgression", "TreeMix", "migration_edges.tsv"),
  type = "treemix"
)
plot_introgression(treemix_edges)
```

ADMIXTOOLS2 qpGraph edge data frames can be written as a table and imported the
same way.

```{r}
qpgraph_edges <- import_introgression(
  ggpop_extdata("introgression", "ADMIXTOOLS2", "qpgraph_edges.tsv"),
  type = "qpgraph"
)
head(qpgraph_edges)
```

## Layered path

Use `ggpop()` plus `geom_introgression()` when composing with additional ggplot
layers.

```{r, fig.alt = "Layered introgression plot using ggpop plus geom_introgression. D statistic windows are drawn as chromosome-wise Manhattan-like points over the genome."}
intro |>
  ggpop() +
  geom_introgression(stat = "D")
```
