---
title: "LD decay"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{LD decay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggPopi)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggPopi", mustWork = TRUE)
}
```

`ggpop` imports LD decay summaries into a typed `ggpop_ld_decay` object. The
module supports PopLDdecay `*.stat.gz` summaries and PLINK pairwise LD tables
summarized into distance bins. PopLDdecay summaries can be left as-is or
collapsed again with `method = "MeanBin"`, `MedianBin`, or
`PercentileBin`.

## API summary

| Task | API | Notes |
|---|---|---|
| Import PopLDdecay summaries | `import_ld_decay(dir, type = "poplddecay")` | Reads `*.stat.gz` and `*.bin.gz` files directly |
| Import PLINK LD pairs | `import_ld_decay(file, type = "plink")` | Summarizes pairwise LD into distance bins |
| Direct plot | `plot_ld_decay(data, style = "point")` | Returns a `ggplot` object |
| Layered plot | `ggpop(data) + geom_ld_decay(...)` | Tidy ggplot extension path |
| Plot style | `style = "point"` / `"line"` | Point summaries or continuous decay curves |
| Measure view | `measure = "r2"` / `"D"` / `"both"` | `D'` needs PopLDdecay D output |
| Population labels | `pop_group` | Uses the package-wide two-column `sample pop` file |

## Import PopLDdecay results

The bundled example contains a PopLDdecay `final_ld.stat.gz` result file. The
importer standardizes `Dist` to `dist`, `Mean_r^2` to `r2`, and `NumberPairs`
to `n_pairs`.

```{r}
ld_dir <- ggpop_extdata("ld_decay", "PopLDdecay")
ld_decay <- import_ld_decay(ld_dir, type = "poplddecay")
class(ld_decay)
head(ld_decay)
```

Population grouping follows the same `pop_group.txt` convention used by PCA and
admixture. The LD decay file label is stored as `sample_id`; if a matching
`sample` appears in `pop_group`, the corresponding `pop` value is used for
colouring. When multiple sample-level summaries map to populations, the plot
layer draws population-level summaries for both point and line styles.

The importer also accepts a single file path, and the `method` argument can be
used to re-bin PopLDdecay summaries with the same mean-bin, median, or
percentile behavior used by the legacy plotting scripts.

```{r}
ld_group_dir <- ggpop_extdata("ld_decay", "PopLDdecay_grouped")
ld_grouped <- import_ld_decay(
  ld_group_dir,
  pop_group = ggpop_extdata("pop_group.txt"),
  type = "poplddecay"
)
unique(ld_grouped$pop)
```

## Point style

The point style follows the common LD summary plot: pairwise distance in Kb on
the x-axis and mean LD \(r^2\) on the y-axis, with population labels mapped to
colour.

```{r, fig.alt = "LD decay point plot. Pairwise distance in kilobases is on the x-axis and mean LD r squared is on the y-axis, with points coloured by population label."}
plot_ld_decay(
  ld_grouped,
  style = "point"
)
```

## Line style

The line style uses the same data and population colour mapping, but draws a
continuous decay curve.

```{r, fig.alt = "LD decay line plot. Pairwise distance in kilobases is on the x-axis and mean LD r squared is on the y-axis, with a continuous curve showing the decay pattern."}
plot_ld_decay(
  ld_grouped,
  style = "line"
)
```

## Layered path

Use `ggpop()` plus `geom_ld_decay()` when composing with other ggplot layers.

```{r, fig.alt = "Layered LD decay point plot using ggpop plus geom_ld_decay. Pairwise distance in kilobases is on the x-axis and mean LD r squared is on the y-axis."}
ld_grouped |>
  ggpop() +
  geom_ld_decay(style = "point")
```
