---
title: "Ne history"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Ne history}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggPopi)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggPopi", mustWork = TRUE)
}
```

`ggpop` imports effective population size histories into a typed
`ggpop_ne_history` object. The module supports PSMC, MSMC2, SMC++, and
Stairway Plot 2 outputs.

Raw VCF and `pop_group.txt` metadata are upstream inputs for these external
demographic inference tools. `ggPopi` reads their output tables; it does not
estimate Ne histories directly from VCF.

## API summary

| Task | API | Notes |
|---|---|---|
| Import PSMC | `import_ne_history(file, type = "psmc")` | Scaled by default; pass `mutation_rate` for absolute values |
| Import MSMC2 | `import_ne_history(file, type = "msmc2")` | Reads final output with time boundaries and lambda columns |
| Import SMC++ | `import_ne_history(file, type = "smcpp")` | Reads `smc++ plot --csv` output (`label,x,y,plot_type,plot_num`) |
| Import Stairway Plot 2 | `import_ne_history(file, type = "stairway")` | Reads `.summary` output with `year`, `Ne_median`, and percentile columns |
| Direct plot | `plot_ne_history(data)` | Returns a `ggplot` object |
| Layered plot | `ggpop(data) + geom_ne_history()` | Tidy ggplot extension path |

## Plot conventions

`plot_ne_history()` defaults to the plotting style used most often by each
upstream workflow. SMC++ fitted histories are drawn as curves. PSMC, MSMC2, and
Stairway Plot 2 histories are drawn as step curves because their native outputs
represent piecewise time intervals. Time and Ne axes are log-scaled by default,
matching the usual demographic-history presentation; pass `log_x = FALSE` or
`log_y = FALSE` for linear axes.

## SMC++ curves

```{r}
smcpp <- import_ne_history(
  ggpop_extdata("ne_history", "SMC++", "model.csv"),
  type = "smcpp"
)
class(smcpp)
head(smcpp)
```

```{r, fig.alt = "Effective population size history line plot. Time before present is on the x-axis and effective population size is on the y-axis, with separate curves for two populations."}
plot_ne_history(smcpp)
```

## Stairway Plot 2 intervals

Stairway Plot 2 summaries often contain lower and upper interval estimates.
When `ne_lower` and `ne_upper` are available, `plot_ne_history()` draws a light
confidence ribbon around the step curve, matching the usual Stairway Plot
summary display of an estimate with interval bounds.

```{r, fig.alt = "Effective population size history with a confidence ribbon. The median Ne curve is shown through time and a translucent band marks lower and upper estimates."}
stairway <- import_ne_history(
  ggpop_extdata("ne_history", "StairwayPlot2", "summary.txt"),
  type = "stairway",
  sample_id = "Stairway"
)
plot_ne_history(stairway, ci = TRUE)
```

## PSMC and MSMC2 scaling

PSMC and MSMC2 outputs are scaled unless a mutation rate is supplied. Passing
`mutation_rate` converts to absolute Ne and time; `generation_time` can then
convert generations to years. Their default plot style is a step curve, which
matches the interval-based output used by `psmc_plot.pl` and MSMC plotting
helpers.

```{r}
psmc_scaled <- import_ne_history(
  ggpop_extdata("ne_history", "PSMC", "sample.psmc"),
  type = "psmc",
  sample_id = "PSMC"
)
unique(psmc_scaled$scale)

msmc_absolute <- import_ne_history(
  ggpop_extdata("ne_history", "MSMC2", "final.txt"),
  type = "msmc2",
  sample_id = "MSMC2",
  mutation_rate = 1e-8
)
unique(msmc_absolute$scale)
```

```{r, fig.alt = "MSMC2 effective population size history as a step curve. Time before present is log-scaled on the x-axis and effective population size is on the y-axis."}
plot_ne_history(msmc_absolute)
```

## Layered path

Use `ggpop()` plus `geom_ne_history()` when composing with additional ggplot
layers.

```{r, fig.alt = "Layered Ne history plot using ggpop plus geom_ne_history. Population-specific SMC++ curves are drawn on log-scaled time and Ne axes."}
smcpp |>
  ggpop() +
  geom_ne_history()
```
