---
title: "PCA plots"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{PCA plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggpop)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggpop", mustWork = TRUE)
}
```

`ggpop` reads PLINK and GCTA PCA output into a typed `ggpop_pca` object.
The same object supports `plot_pca()` and the layered `ggpop() + geom_pca()` style.

## API summary

| Task | API | Notes |
|---|---|---|
| Import PLINK/GCTA PCA | `import_pca(file, type = ..., eigenval = NULL, pop_group = NULL)` | Returns `ggpop_pca` |
| Direct plot | `plot_pca(data, pc_x = 1, pc_y = 2)` | Returns `ggplot` |
| Layered plot | `ggpop(data) + geom_pca(pc_x = 1, pc_y = 2)` | Tidy ggplot extension path |
| Compute PCA | `compute_pca(genotype, method = "flashpca")` | Requires `flashpcaR` |

## GCTA PCA

```{r}
pca <- import_pca(
  ggpop_extdata("pca", "gcta.eigenvec"),
  type = "gcta",
  eigenval = ggpop_extdata("pca", "gcta.eigenval"),
  pop_group = ggpop_extdata("pop_group.txt")
)
```

```{r}
plot_pca(pca, title = "GCTA PCA")
```

```{r}
ggpop(pca) + geom_pca(pc_x = 1, pc_y = 3)
```

When the imported object has a `pop` column, both plotting routes map population groups to the unified ggpop discrete colour scale.

```{r}
ggpop(pca) +
  geom_pca(pc_x = 1, pc_y = 2, palette = "population")
```

## PC labels

The axis labels can carry variance explained when eigenvalues are available.

```{r}
plot_pca(pca, pc_x = 1, pc_y = 2)
```

## Optional flashpcaR computation

`compute_pca(method = "flashpca")` is available when `flashpcaR` is installed.
It is intentionally not forced during vignette build because it depends on a genotype object.

```{r eval=FALSE}
compute_pca(genotype, method = "flashpca")
```
