---
title: "Selective sweep scans"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Selective sweep scans}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggpop)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggpop", mustWork = TRUE)
}
```

`ggpop` imports selection scan outputs into a typed `ggpop_selection` object.
The module supports common selscan normalized outputs and XPCLR window tables.

## API summary

| Task | API | Notes |
|---|---|---|
| Import selscan results | `import_selection(dir, type = "selscan")` | Auto-discovers iHS, nSL, iHH12, XP-EHH, and XP-nSL norm files |
| Import XPCLR windows | `import_selection(dir, type = "xpclr")` | Uses window midpoint as `pos` and keeps `start` / `end` |
| Direct plot | `plot_selection(data, stat = ..., chr = ...)` | Returns a `ggplot` object |
| Layered plot | `ggpop(data) + geom_selection(...)` | Tidy ggplot extension path |
| Region filter | `chr`, `start`, `end` | Keeps points or windows overlapping the region |
| Plot style | `style = "auto"` / `"single"` / `"manhattan"` | Genome-wide calls default to Manhattan-like; local calls default to single-region |
| Thresholds | `threshold = 2` or `threshold = 0.95, threshold_type = "quantile"` | Fixed score cutoffs or filtered-data quantiles |
| Score sign | `value = "signed"` / `"absolute"` | Signed z-score-like views or absolute outlier magnitude |

## Import selscan results

The bundled selscan example contains normalized iHS, nSL, iHH12, XP-EHH, and
XP-nSL outputs. Directory imports auto-discover supported result files.

```{r}
selscan_dir <- ggpop_extdata("selective_sweep", "selscan")
selscan <- import_selection(selscan_dir, type = "selscan")
class(selscan)
unique(selscan$stat)
```

You can also import selected files explicitly. Relative paths are resolved
inside `dir`.

```{r}
selscan_chr1 <- import_selection(
  selscan_dir,
  ihs = "chr1.ihs.out.100bins.norm",
  nsl = "chr1.nsl.out.100bins.norm",
  xpehh = "chr1.xpehh.out.norm",
  xpnsl = "chr1.xpnsl.out.norm",
  type = "selscan"
)
unique(selscan_chr1$stat)
```

## Plot scan statistics

`plot_selection()` stacks selected statistics vertically and keeps genomic
position aligned on the x-axis. Region filters work on points for selscan and
on overlapping windows for XPCLR.
By default, genome-wide calls use a Manhattan-like chromosome axis. Calls with
`chr`, `start`, or `end` use the single-region style, where position is shown
directly in megabases.

```{r, fig.alt = "Faceted scatter plot of four selscan statistics on chromosome 1. iHS, nSL, XP-EHH, and XP-nSL are stacked vertically with genomic position in megabases on the shared x-axis."}
plot_selection(
  selscan_chr1,
  stat = c("ihs", "nsl", "xpehh", "xpnsl"),
  chr = "1"
)
```

The layered API follows the same grammar as the other modules:

```{r, fig.alt = "Layered ggplot selection scan. The ggpop object supplies imported selscan data, and geom_selection draws iHS values for chromosome 1 with horizontal threshold lines."}
selscan_chr1 |>
  ggpop() +
  geom_selection(stat = "ihs", chr = "1", threshold = 2)
```

Use `value = "absolute"` when the direction of the statistic is less important
than the magnitude. Quantile thresholds are computed after the statistic,
chromosome, and region filters are applied.

```{r, fig.alt = "Absolute iHS selection scan on chromosome 1. The y-axis shows absolute selection score magnitude, with a dashed horizontal line at the 95th percentile of the filtered chromosome 1 iHS values."}
plot_selection(
  selscan_chr1,
  stat = "ihs",
  chr = "1",
  value = "absolute",
  threshold = 0.95,
  threshold_type = "quantile"
)
```

## Manhattan-like genome axis

For genome-wide scan summaries, it is often useful to lay chromosomes end to
end like a Manhattan plot. Fine-scale structure is compressed, but this view is
good for seeing chromosome-scale shifts and scan outliers across the genome.

```{r}
selscan_selected <- import_selection(
  selscan_dir,
  ihs1 = "chr1.ihs.out.100bins.norm",
  ihs2 = "chr2.ihs.out.100bins.norm",
  ihs3 = "chr3.ihs.out.100bins.norm",
  nsl1 = "chr1.nsl.out.100bins.norm",
  nsl2 = "chr2.nsl.out.100bins.norm",
  nsl3 = "chr3.nsl.out.100bins.norm",
  xpehh1 = "chr1.xpehh.out.norm",
  xpehh2 = "chr2.xpehh.out.norm",
  xpehh3 = "chr3.xpehh.out.norm",
  xpnsl1 = "chr1.xpnsl.out.norm",
  xpnsl2 = "chr2.xpnsl.out.norm",
  xpnsl3 = "chr3.xpnsl.out.norm",
  type = "selscan"
)

finite_scan <- selscan_selected[is.finite(selscan_selected$value), ]
```

Show absolute iHS on its own scale:

```{r, fig.alt = "Manhattan-like selection scan for iHS. Chromosomes 1, 2, and 3 are placed end to end on a continuous genome position axis in megabases, and absolute iHS is plotted as small points coloured by chromosome."}
plot_selection(
  finite_scan,
  stat = "ihs",
  style = "manhattan",
  value = "absolute",
  threshold = 0.95,
  threshold_type = "quantile"
)
```

Plot normalized nSL, XP-EHH, and XP-nSL as z-score-like panels:

```{r, fig.alt = "Manhattan-like faceted selection scan. nSL, XP-EHH, and XP-nSL are shown in stacked panels over a continuous genome position axis in megabases, with points coloured by chromosome."}
plot_selection(
  finite_scan,
  stat = c("nsl", "xpehh", "xpnsl"),
  style = "manhattan"
)
```

## Import XPCLR windows

XPCLR outputs are window-based. The importer keeps the original `start` and
`end` columns, computes `pos` as the midpoint, and prefers `xpclr_norm` when it
is available.

```{r}
xpclr_dir <- ggpop_extdata("selective_sweep", "xpclr")
xpclr <- import_selection(xpclr_dir, type = "xpclr")
class(xpclr)
unique(xpclr$stat)
head(xpclr[c("chr", "start", "end", "pos", "value")])
```

```{r, fig.alt = "XPCLR Manhattan-like genome scan. XPCLR window midpoints are placed along a chromosome axis, with normalized XPCLR score on the y-axis and points coloured by chromosome."}
plot_selection(
  xpclr,
  stat = "xpclr",
  style = "manhattan",
  threshold = 0.95,
  threshold_type = "quantile"
)
```

Use `geom_selection()` when the scan should be composed with additional ggplot
layers or annotations.
