---
title: "Population genomics statistics"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Population genomics statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggpop)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggpop", mustWork = TRUE)
}
```

`ggpop` imports windowed population genomics statistics into a typed
`ggpop_stats` object. The current module supports pixy and vcftools-style
outputs for common summaries including FST, pi, Tajima's D, Dxy, and
Watterson's theta.

## API summary

| Task | API | Notes |
|---|---|---|
| Import a result directory | `import_stats(dir, type = "pixy")` | Auto-discovers supported suffixes |
| Import selected files | `import_stats(dir, pi = "...", fst = "...")` | Relative paths resolve inside `dir` |
| Direct plot | `plot_stats(data, stat = ..., chr = ...)` | Returns a `ggplot` object |
| Layered plot | `ggpop(data) + geom_stats(...)` | Tidy ggplot extension path |
| Region filter | `chr`, `start`, `end` | Keeps windows overlapping the region |

## Import pixy results

The bundled pixy example is a dense chr2L windowed data set. It contains 4,000
tidy rows across five statistics: 500 pairwise Dxy windows, 500 pairwise FST
windows, and 1,000 within-population windows each for pi, Tajima's D, and
Watterson's theta. Windows are 100 bp wide and span positions 1--50,000.

```{r}
pixy_dir <- ggpop_extdata("Population_genomics_statistics", "pixy")
stats <- import_stats(pixy_dir, type = "pixy")
class(stats)
unique(stats$stat)
range(stats$start)
```

You can also import selected files explicitly:

```{r}
selected <- import_stats(
  pixy_dir,
  pi = "pixy_pi.txt",
  fst = "pixy_fst.txt",
  tajima = "pixy_tajima_d.txt",
  type = "pixy"
)
unique(selected$stat)
```

## Import vcftools results

vcftools results use a different column naming convention, but enter the same
typed plotting object. Supported examples include `.windowed.pi`,
`.windowed.weir.fst`, and `.Tajima.D` files.

```{r}
vcftools_dir <- ggpop_extdata("Population_genomics_statistics", "vcftools")
vcftools_stats <- import_stats(vcftools_dir, type = "vcftools")
class(vcftools_stats)
unique(vcftools_stats$stat)
```

Explicit file selection works the same way:

```{r}
vcftools_selected <- import_stats(
  vcftools_dir,
  pi = "vcftools.windowed.pi",
  fst = "vcftools.windowed.weir.fst",
  tajima = "vcftools.Tajima.D",
  type = "vcftools"
)
unique(vcftools_selected$stat)
```

## Plot all statistics

By default, `stat = "all"` stacks the selected statistics vertically and keeps
the x-axis aligned. This mirrors the usual pixy plotting layout for comparing
windowed summaries across the same genomic coordinate system. For the bundled
example, all panels use chr2L 100 bp windows across the first 50 kb.

```{r, fig.alt = "Faceted line plot of population genomics statistics across chr2L. Five statistic panels are stacked vertically for Dxy, FST, pi, Tajima's D, and Watterson's theta. Genomic position in megabases is on the shared x-axis from 1 to 50,000 bases, and each panel has its own statistic value scale."}
plot_stats(stats, stat = "all", chr = "chr2L")
```

The same plotting path works for vcftools imports:

```{r, fig.alt = "Faceted line plot from vcftools results. FST, pi, and Tajima's D are stacked vertically, genomic position in megabases is on the x-axis, and each panel has its own statistic value scale."}
plot_stats(vcftools_stats, stat = "all", chr = "chr2L")
```

## Select statistics and regions

Use `stat` to choose one or more summaries, `chr` to select chromosomes, and
`start` / `end` to focus on windows overlapping a region.

```{r, fig.alt = "Two-panel line plot. FST and pi are stacked vertically for chromosome chr2L using 100 bp windows, with genomic position in megabases on the shared x-axis and each statistic value on its own y-axis."}
plot_stats(stats, stat = c("fst", "pi"), chr = "chr2L")
```

```{r, fig.alt = "Regional line plot of population genomics statistics. Windows on chromosome chr2L overlapping bases 1 to 20,000 are shown, focusing all statistic panels on the first two-fifths of the bundled 50 kb example region."}
plot_stats(stats, chr = "chr2L", start = 1, end = 20000)
```

Multiple chromosomes can be selected. When more than one chromosome is shown,
the default plot uses points to avoid implying continuous genomic distance
between chromosomes. The bundled pixy example contains only chr2L, so the
multi-chromosome example below uses the vcftools import if multiple chromosomes
are present in user data.

```{r eval=FALSE}
plot_stats(stats, chr = c("chr1", "chr2"))
```

## Layered workflow

The layered API follows the same grammar as the other modules:

```{r, fig.alt = "Layered line plot of FST and pi. The ggpop object supplies imported population statistics and geom_stats draws two vertically stacked statistic panels for chromosome chr2L."}
stats |>
  ggpop() +
  geom_stats(stat = c("fst", "pi"), chr = "chr2L")
```

The module uses ggpop's shared theme, font sizing, and discrete palette
interfaces. Use `palette`, `base_size`, and `base_family` the same way as the
GWAS and PCA modules.
