---
title: "Population genomics statistics"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Population genomics statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggpop)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggpop", mustWork = TRUE)
}
```

`ggpop` imports windowed population genomics statistics into a typed
`ggpop_stats` object. The current module supports pixy and vcftools-style
outputs for common summaries including FST, pi, Tajima's D, Dxy, and
Watterson's theta.

## API summary

| Task | API | Notes |
|---|---|---|
| Import a result directory | `import_stats(dir, type = "pixy")` | Auto-discovers supported suffixes |
| Import selected files | `import_stats(dir, pi = "...", fst = "...")` | Relative paths resolve inside `dir` |
| Direct plot | `plot_stats(data, stat = ..., chr = ...)` | Returns a `ggplot` object |
| Layered plot | `ggpop(data) + geom_stats(...)` | Tidy ggplot extension path |
| Region filter | `chr`, `start`, `end` | Keeps windows overlapping the region |

## Import pixy results

```{r}
pixy_dir <- ggpop_extdata("Population_genomics_statistics", "pixy")
stats <- import_stats(pixy_dir, type = "pixy")
class(stats)
unique(stats$stat)
```

You can also import selected files explicitly:

```{r}
selected <- import_stats(
  pixy_dir,
  pi = "pixy_pi.txt",
  fst = "pixy_fst.txt",
  tajima = "pixy_tajima_d.txt",
  type = "pixy"
)
unique(selected$stat)
```

## Plot all statistics

By default, `stat = "all"` stacks the selected statistics vertically and keeps
the x-axis aligned. This mirrors the usual pixy plotting layout for comparing
windowed summaries across the same genomic coordinate system.

```{r, fig.alt = "Faceted line plot of population genomics statistics. Statistic names define vertically stacked panels, genomic position in megabases is on the x-axis, and statistic value is on the y-axis. The shared x-axis lets FST, pi, Tajima's D, Dxy, and Watterson's theta be compared across the same chromosome windows."}
plot_stats(stats, stat = "all", chr = "chr2L")
```

## Select statistics and regions

Use `stat` to choose one or more summaries, `chr` to select chromosomes, and
`start` / `end` to focus on windows overlapping a region.

```{r, fig.alt = "Two-panel line plot. FST and pi are stacked vertically for chromosome chr2L, with genomic position in megabases on the x-axis and each statistic value on its own y-axis."}
plot_stats(stats, stat = c("fst", "pi"), chr = "chr2L")
```

```{r, fig.alt = "Regional line plot of population genomics statistics. Windows on chromosome chr2L overlapping the first 20,000 bases are shown, focusing the x-axis on a small genomic interval."}
plot_stats(stats, chr = "chr2L", start = 1, end = 20000)
```

Multiple chromosomes can be selected. When more than one chromosome is shown,
the default plot uses points to avoid implying continuous genomic distance
between chromosomes.

```{r, fig.alt = "Faceted point plot of population genomics statistics across two chromosomes. Statistic panels are stacked vertically, point position shows genomic window midpoint, and colour separates chromosome categories."}
plot_stats(stats, chr = c("chr2L", "chr3L"))
```

## Layered workflow

The layered API follows the same grammar as the other modules:

```{r, fig.alt = "Layered line plot of FST and pi. The ggpop object supplies imported population statistics and geom_stats draws two vertically stacked statistic panels for chromosome chr2L."}
stats |>
  ggpop() +
  geom_stats(stat = c("fst", "pi"), chr = "chr2L")
```

The module uses ggpop's shared theme, font sizing, and discrete palette
interfaces. Use `palette`, `base_size`, and `base_family` the same way as the
GWAS and PCA modules.
