---
title: "Getting Started with ggpop"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Getting Started with ggpop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggpop)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggpop", mustWork = TRUE)
}
```

`ggpop` is a ggplot2 extension package for population genetics workflows.
The package keeps each module in the same tidy shape:

- `import_*()` functions create typed S3 objects.
- `plot_*()` functions return full `ggplot` objects.
- `ggpop()` + `geom_*()` build layered ggplot extensions.
- Advanced compatibility helpers remain exported for users who need original-package behavior.

## Module API map

| Module | Import | Direct plot | ggplot extension path | Advanced / compatibility |
|---|---|---|---|---|
| GWAS Manhattan | `import_gwas()` | `plot_manha()` | `ggpop() + geom_manha()` | internal fastman-style layout |
| GWAS Q-Q | `import_gwas()` | `plot_qq()` | `ggpop() + ggpop::geom_qq()` | internal fastqq-style layout |
| PCA | `import_pca()` / `compute_pca()` | `plot_pca()` | `ggpop() + geom_pca()` | `compute_pca(method = "flashpca")` |
| Admixture | `import_admix()` / `import_admixture()` | `plot_admix()` | `ggpop() + geom_admix()` | see compatibility article |
| Population statistics | `import_stats()` | `plot_stats()` | `ggpop() + geom_stats()` | pixy and vcftools summaries |

## Core pattern

```{r}
gwas <- import_gwas(ggpop_extdata("gwas", "gcta.mlma"), type = "gcta")
pca <- import_pca(
  ggpop_extdata("pca", "gcta.eigenvec"),
  type = "gcta",
  eigenval = ggpop_extdata("pca", "gcta.eigenval"),
  pop_group = ggpop_extdata("pop_group.txt")
)
admix <- import_admix(
  ggpop_extdata("admixture"),
  type = "admixture",
  ind = ggpop_extdata("snp", "finalsnp_ld.fam"),
  pop_group = ggpop_extdata("pop_group.txt")
)
stats <- import_stats(
  ggpop_extdata("Population_genomics_statistics", "pixy"),
  type = "pixy"
)
```

Each importer returns a typed object:

```{r}
class(gwas)
class(pca)
class(admix)
class(stats)
```

## Tidy plotting style

Every module has two user-facing plotting paths. Use the direct `plot_*()`
function when you want the reference plot immediately, or use `ggpop()` plus
the module `geom_*()` when you want to compose with other ggplot layers.

The direct path:

```{r, fig.alt = "Manhattan plot. Chromosomes are arranged along the x-axis and minus log10 p-values are on the y-axis. Points form chromosome-specific bands with horizontal reference lines marking suggestive and genome-wide significance thresholds."}
gwas |>
  plot_manha(title = "GCTA Manhattan")
```

The ggplot extension path:

```{r, fig.alt = "Manhattan plot from the layered ggplot extension path. Chromosomes are arranged along the x-axis and minus log10 p-values are on the y-axis, matching the direct plot_manha visual style."}
gwas |>
  ggpop() +
  geom_manha()
```

The same pattern applies across modules:

```{r, fig.alt = c("Q-Q scatter plot. Expected minus log10 p-values are on the x-axis and observed minus log10 p-values are on the y-axis, with points compared against a diagonal reference line.", "Scatter chart. Principal component 1 is on the x-axis and principal component 2 is on the y-axis, with point colour encoding population groups.", "Stacked bar chart. Individuals are arranged along the x-axis and ancestry proportions are stacked within each bar for K equals 3.")}
gwas |> ggpop() + ggpop::geom_qq()
pca |> ggpop() + geom_pca()
admix |> ggpop() + geom_admix(k = 3, order_group = TRUE)
```

## Population groups and discrete colours

Population grouping uses a simple two-column file:

```{r, fig.alt = "Two-column population group table imported from pop_group.txt with sample and pop columns."}
head(import_pop_group(ggpop_extdata("pop_group.txt")))
```

The same file drives PCA colours and admixture group labels:

```{r, fig.alt = c("Scatter chart. Principal component 1 is on the x-axis and principal component 2 is on the y-axis, with point colour encoding imported population groups.", "Stacked bar chart. Individuals are arranged along the x-axis and ancestry proportions are stacked within each bar, with population group labels used for ordering.")}
pca |> plot_pca(title = "GCTA PCA by population")
admix |> plot_admix(k = 3, order_group = TRUE, show_group_labels = TRUE)
```

All categorical colours use a unified discrete palette entry:

```{r, fig.alt = "Two discrete palette outputs. The first vector lists four population colours, and the second vector lists eight admixture cluster colours for categorical encodings."}
ggpop_palette(4, "population")
ggpop_palette(8, "admixture")
```

## What to use

- Use `plot_manha()` and `ggpop() + geom_manha()` for Manhattan plots.
- Use `plot_qq()` and `ggpop() + ggpop::geom_qq()` for Q-Q plots.
- Use `plot_pca()` and `ggpop() + geom_pca()` for PCA plots.
- Use `plot_admix()` and `ggpop() + geom_admix()` for admixture plots.
- Use `plot_stats()` and `ggpop() + geom_stats()` for windowed population statistics.
- Treat the direct `plot_*()` functions as the reference style; `geom_*()` is the same look inside a ggplot composition.
- Use the compatibility article only when you need original `pophelper` workflows.
