---
title: "Getting Started with ggpop"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Getting Started with ggpop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggpop)

ggpop_extdata <- function(...) {
  system.file("extdata", ..., package = "ggpop", mustWork = TRUE)
}
```

`ggpop` is a ggplot2 extension package for population genetics workflows.
The package keeps each module in the same tidy shape:

- `import_*()` functions create typed S3 objects.
- `plot_*()` functions return full `ggplot` objects.
- `ggpop()` + `geom_*()` build layered ggplot extensions.
- Advanced compatibility helpers remain exported for users who need original-package behavior.

## Module API map

| Module | Import | Direct plot | ggplot extension path | Advanced / compatibility |
|---|---|---|---|---|
| GWAS Manhattan | `import_gwas()` | `plot_manha()` | `ggpop() + geom_manha()` | internal fastman-style layout |
| GWAS Q-Q | `import_gwas()` | `plot_qq()` | `ggpop() + ggpop::geom_qq()` | internal fastqq-style layout |
| PCA | `import_pca()` / `compute_pca()` | `plot_pca()` | `ggpop() + geom_pca()` | `compute_pca(method = "flashpca")` |
| Admixture | `import_admix()` | `plot_admix()` | `ggpop() + geom_admix()` | see compatibility article |
| Population statistics | `import_stats()` | `plot_stats()` | `ggpop() + geom_stats()` | pixy and vcftools summaries |
| LD decay | `import_ld_decay()` | `plot_ld_decay()` | `ggpop() + geom_ld_decay()` | PopLDdecay and PLINK summaries |
| Selective sweeps | `import_selection()` | `plot_selection()` | `ggpop() + geom_selection()` | selscan and XPCLR scans |
| Introgression | `import_introgression()` | `plot_introgression()` | `ggpop() + geom_introgression()` | Dsuite, genomics_general, TreeMix, and qpGraph summaries |

## Core pattern

```{r}
gwas <- import_gwas(ggpop_extdata("gwas", "gcta.mlma"), type = "gcta")
pca <- import_pca(
  ggpop_extdata("pca", "gcta.eigenvec"),
  type = "gcta",
  eigenval = ggpop_extdata("pca", "gcta.eigenval"),
  pop_group = ggpop_extdata("pop_group.txt")
)
admix <- import_admix(
  ggpop_extdata("admixture"),
  type = "admixture",
  ind = ggpop_extdata("snp", "finalsnp_ld.fam"),
  pop_group = ggpop_extdata("pop_group.txt")
)
stats <- import_stats(
  ggpop_extdata("Population_genomics_statistics", "pixy"),
  type = "pixy"
)
ld_decay <- import_ld_decay(
  ggpop_extdata("ld_decay", "PopLDdecay"),
  type = "poplddecay"
)
selscan_chr1 <- import_selection(
  ggpop_extdata("selective_sweep", "selscan"),
  ihs = "chr1.ihs.out.100bins.norm",
  nsl = "chr1.nsl.out.100bins.norm",
  xpehh = "chr1.xpehh.out.norm",
  xpnsl = "chr1.xpnsl.out.norm",
  type = "selscan"
)
introgression <- import_introgression(
  ggpop_extdata("introgression", "genomics_general"),
  type = "genomics_general"
)
```

Each importer returns a typed object:

```{r}
class(gwas)
class(pca)
class(admix)
class(stats)
class(ld_decay)
class(selscan_chr1)
class(introgression)
```

## Tidy plotting style

Every module has two user-facing plotting paths. Use the direct `plot_*()`
function when you want the reference plot immediately, or use `ggpop()` plus
the module `geom_*()` when you want to compose with other ggplot layers.

The direct path:

```{r, fig.alt = "Manhattan plot. Chromosomes are arranged along the x-axis and minus log10 p-values are on the y-axis. Points form chromosome-specific bands with horizontal reference lines marking suggestive and genome-wide significance thresholds."}
gwas |>
  plot_manha()
```

The ggplot extension path:

```{r, fig.alt = "Manhattan plot from the layered ggplot extension path. Chromosomes are arranged along the x-axis and minus log10 p-values are on the y-axis, matching the direct plot_manha visual style."}
gwas |>
  ggpop() +
  geom_manha()
```

The same pattern applies across modules:

```{r, fig.alt = c("Q-Q scatter plot. Expected minus log10 p-values are on the x-axis and observed minus log10 p-values are on the y-axis, with points compared against a diagonal reference line.", "Scatter chart. Principal component 1 is on the x-axis and principal component 2 is on the y-axis, with point colour encoding population groups.", "Stacked bar chart. Individuals are arranged along the x-axis and ancestry proportions are stacked within each bar for K equals 3.")}
gwas |> ggpop() + ggpop::geom_qq()
pca |> ggpop() + geom_pca()
admix |> ggpop() + geom_admix(k = 3, order_group = TRUE)
```

## Population groups and discrete colours

Population grouping uses a simple two-column file:

```{r, fig.alt = "Two-column population group table imported from pop_group.txt with sample and pop columns."}
head(import_pop_group(ggpop_extdata("pop_group.txt")))
```

The same file drives PCA colours and admixture group labels:

```{r, fig.alt = c("Scatter chart. Principal component 1 is on the x-axis and principal component 2 is on the y-axis, with point colour encoding imported population groups.", "Stacked bar chart. Individuals are arranged along the x-axis and ancestry proportions are stacked within each bar, with population group labels used for ordering.")}
pca |> plot_pca()
admix |> plot_admix(k = 3, order_group = TRUE, show_group_labels = TRUE)
```

All categorical colours use a unified discrete palette entry:

```{r, fig.alt = "Two discrete palette outputs. The first vector lists four population colours, and the second vector lists eight admixture cluster colours for categorical encodings."}
ggpop_palette(4, "population")
ggpop_palette(8, "admixture")
```

## Population statistics

The statistics module uses the same tidy pattern for windowed summaries from
pixy or vcftools outputs:

```{r, fig.alt = "Faceted line plot of population genomics statistics on chr2L. Dxy, FST, pi, Tajima's D, and Watterson's theta are stacked vertically, genomic position in megabases is on the shared x-axis, and each panel uses its own statistic value scale."}
stats |>
  plot_stats(stat = "all", chr = "chr2L")
```

The layered path filters facets to the requested statistics:

```{r, fig.alt = "Layered two-panel line plot of population genomics statistics. The ggpop object supplies imported pixy statistics and geom_stats draws only the selected FST and pi facets for chr2L."}
stats |>
  ggpop() +
  geom_stats(stat = c("fst", "pi"), chr = "chr2L")
```

## LD decay

LD decay summaries use the same direct and layered plotting shape. PopLDdecay
`*.stat.gz` files are imported directly, while PLINK pairwise LD files can be
summarized into distance bins. Population labels follow the package-wide
`pop_group.txt` convention when file labels need to be mapped to groups.

```{r, fig.alt = "LD decay point plot. Pairwise distance in kilobases is on the x-axis and mean LD r squared is on the y-axis, with points coloured by population label."}
ld_decay |>
  plot_ld_decay(style = "point")
```

The same data can be drawn as a curve:

```{r, fig.alt = "LD decay line plot. Pairwise distance in kilobases is on the x-axis and mean LD r squared is on the y-axis, with a continuous curve showing the decay pattern."}
ld_decay |>
  plot_ld_decay(style = "line")
```

## Selective sweep scans

Selection scan outputs use the same direct and layered plotting shape. The
direct plot can show signed values or absolute score magnitude, and thresholds
can be fixed values or filtered-data quantiles. Genome-wide calls default to a
Manhattan-like chromosome axis; calls with `chr`, `start`, or `end` default to a
single-region view.

```{r, fig.alt = "Faceted selection scan on chromosome 1. iHS, nSL, XP-EHH, and XP-nSL are stacked vertically with genomic position in megabases on the x-axis."}
selscan_chr1 |>
  plot_selection(
    stat = c("ihs", "nsl", "xpehh", "xpnsl"),
    chr = "1"
  )
```

## Introgression

Introgression summaries use the same direct and layered plotting shape.
Windowed Dsuite and genomics_general outputs default to a Manhattan-like
chromosome axis; Dsuite Dtrios summaries use a trio-level dot plot; graph edge
tables use a compact edge diagram.

```{r, fig.alt = "Manhattan-like introgression plot. D and fdM window statistics are shown in stacked panels over chromosomes, with points coloured by chromosome."}
introgression |>
  plot_introgression(stat = c("D", "fdM"))
```

The layered path follows the same grammar:

```{r, fig.alt = "Layered introgression plot using ggpop plus geom_introgression. D statistic windows are shown over the genome with a Manhattan-like chromosome axis."}
introgression |>
  ggpop() +
  geom_introgression(stat = "D")
```

## What to use

- Use `plot_manha()` and `ggpop() + geom_manha()` for Manhattan plots.
- Use `plot_qq()` and `ggpop() + ggpop::geom_qq()` for Q-Q plots.
- Use `plot_pca()` and `ggpop() + geom_pca()` for PCA plots.
- Use `plot_admix()` and `ggpop() + geom_admix()` for admixture plots.
- Use `plot_stats()` and `ggpop() + geom_stats()` for windowed population statistics.
- Use `plot_ld_decay()` and `ggpop() + geom_ld_decay()` for LD decay summaries.
- Use `plot_selection()` and `ggpop() + geom_selection()` for selective sweep scans.
- Use `plot_introgression()` and `ggpop() + geom_introgression()` for introgression summaries.
- Treat the direct `plot_*()` functions as the reference style; `geom_*()` is the same look inside a ggplot composition.
- Use the compatibility article only when you need original `pophelper` workflows.
